Measuring dimension

### 2012-08-29 Dimension/src/dimension.tex
...

\documentclass[10pt]{article}

@@

The logic behind assumption \ref{box-counting-assumption} is detailed in @!. However, this relation is not guaranteed to hold exactly. A common tactic is to assume the relation 
\begin{align}
N_\epsilon = \Phi(\frac{1}{\epsilon})\left (\frac{1}{\epsilon}\right) ^ d
\end{align}

@@

We can change the definition of dimension by introducing a parameter $q$ that determines how much more we care about regions with high probability than about regions with low probability. We define the quantity $I_q$ as an analgoue of the box counting dimension:
\[
I_q(p, \epsilon) = \sum_{c \in C_\epsilon}p(c)^q 
\]
Where $C$ is the covering of the probability distribution in boxes of side length $\epsilon$. The \textit{generalized dimension} of $X$ is then defined as
\[
D_q(p) = \frac{1}{q-1} \lim_{\epsilon \rightarrow 0} \frac{\ln I_q(X, \epsilon)}{\ln \epsilon}
\]

For $D_0$ reduces to the box counting dimension of the support. $D_1$ causes a division by zero, but letting $q \rightarrow 1$ we can formulate 
\[
D_1(p) = lim_{\epsilon \rightarrow 0} \frac{1}{\ln \epsilon} \sum_{c \in C_\epsilon} p(c) \ln p(\epsilon)
\] 
The right part of this is the negative of the entropy of the distribution over $C$ under $p$. In essence $D_1$ measures how the entropy of a covering scales with the accuracy of that covering.

@@

As noted before, we assume that our data is independently sampled from some probability distribution. This may not be the case. In particular time series data\ldots

Read Theiler's EEG paper. Discuss natural measures briefly. Also ergodicity?

@@

Discussion of the running time an an investigation of the effects of reducing the data size.

@@

This is an example of a distribution with multiple intrinsic dimensions. The data is generated by different processes that each create a dimensionality in the data at a particular scale. It's not uncommon to see data with more than two intrinsic dimensions. In the log-log plot produces by the Grassberger-Procaccia estimator, this behavior will produce linear segments for the relevant length scales.

@@

To find the interval limits $r_i$ we proced as we did before with $d_{\max}$ and search for a set of values from the data which minimize the Kolmogorov-Smirnov statistic with the filtered data. We compare each interval with the data filtered to that interval and take the sum of all intervals for a given set as the value to minimize. As searching a for a single value was already expensive in the single-scale case, here we must give up on the idea of checking all possible values and instead search for a suitable set of ranges.

@!NOTE: Try contiguous and non-contiguous scaling intervals.

@@

Let $R$ be the set of distances in the dataset less than $r_{\max}$. The the likelihood of this dataset is
\[
{\cal L}(R|D_t) = \prod_{r \in R} D_t \frac{1}{r} \left (\frac{r}{r_{\max}} \right ) ^ {D_t} 
\] 
Which we must maximize. Maximizing the logarithm of the likelihood is equivalent, and easier.
\begin{align*}
\ln {\cal L}(R|D_t) &= -\sum \ln r + D_t \sum \ln r - D_t |R| \ln r_{\max} + |R| \ln D_t
\end{align*}

@@

\begin{align*}
\frac{\partial \ln{\cal L}(R|D_t)}{\partial D_t} = \sum{\ln r} - |R|\ln r_{\max} + |R|\frac{1}{D_t}
\end{align*}

@@

\begin{align*}
 \sum{\ln r} - |R|\ln r_{\max} + |R|\frac{1}{D_t} &= 0 \\
|R|\frac{1}{D_t} &= - \sum \ln r + |R| \ln r_{\max} \\
D_t &= \frac{|R|}{- \sum \ln r + |R| \ln r_{\max}} 
\end{align*}

Many publications assume that the data $|R|$ are normalized by $r_{\max}$ which allows the cleaner result:
\begin{align*}
D_t &= \frac{|R|}{- \sum \ln r} 
\end{align*}

However, this hides the reliance on $r_{\max}$. As we will be iterating over all possible values for $r_{\max}$, it's easier to include in the estimator explicitly.

@@

We start with the cumulative distribution function as described in the text. We use $r_{i-1} = r_h$ tohelp with the notation.

@@

\begin{align*}
p(d = r \mid \ldots) = {D_i}r^{D_i-1} \frac{1}{\left(\frac{r_i}{r_j}\right)^{D_i} - 1}
\end{align*}

This gives us the following log likelihood function:
\begin{align*}
\ln{\cal L}(R_i|D_i) &= \sum \ln D_i + (D_i - 1)\sum \ln r - \sum \ln \left(\left(\frac{r_i}{r_j}\right)^{D_i} -1\right)\\
&=  |R_i|\ln D_i + (D_i - 1)\sum \ln r - |R_i| \ln \left(\left(\frac{r_i}{r_j}\right)^{D_i} -1\right) 
\end{align*}

Where $R_i$ is the subset of distances in the dataset that fall within the interval $[r_{i-1}, r_i)$.

Taking the derivative:
\begin{align*}
\frac {\partial \ln{\cal L}(R_i|D_i)}{\partial D_i} &=  \frac{|R_i|}{D_i} + \sum \ln r - |R_i| \frac{\left(\frac{r_i}{r_j}\right)^{D_i} \ln \frac{r_i}{r_j}}{\left(\frac{r_i}{r_j}\right)^{D_i} -1}
\end{align*}

@@

\begin{align*}
\frac{|R_i|}{D_i} + \sum \ln r - |R_i| \frac{\left(\frac{r_i}{r_j}\right)^{D_i} \ln \frac{r_i}{r_j}}{\left(\frac{r_i}{r_j}\right)^{D_i} -1} &= 0 \\
\frac{1}{D_i}  - \frac{\left(\frac{r_i}{r_j}\right)^{D_i} \ln \frac{r_i}{r_j}}{\left(\frac{r_i}{r_j}\right)^{D_i} -1} - \frac{-\sum \ln r}{|R_i|} &= 0  
\end{align*}

Wehave little hope of solving this for $D_i$, but as shown in lemma \ref{monotonicity}, the left of this equation is strictly decreasing in the range over which we are interested. This allows us to find the value of $D_i$ by binary search as detailed in the main text.

@@

@@

\begin{lma}
\label{monotonicity}
The function
\[
f(x) = \frac{1}{x} + \frac{a^x \ln(a)}{a^x - 1} + c
\]
is strictly decreasing for $ x \in (0, \infty)$ and $ a \in (0, 1)$.
\end{lma}
\begin{proof}
Taking the derivative gives us.

@@

$f$ is stricly decreasing in the given domain if $f'(x) < 0$ holds. Assume for a contradiction that
\begin{align*}
-\frac{1}{x^2} + \left [ \frac{\sqrt{a^x}\ln a}{a^x - 1} \right ]^2 \geq 0
\end{align*}
Which rewrites to 
\begin{align*}
\frac{1}{x}  &\leq \frac{\sqrt{a^x}\ln a}{a^x - 1}
x  &\geq \frac{a^x - 1}{\sqrt{a^x}\ln a}
\end{align*}
We know that $x > 0$ so that 
\begin{align*}
0  &> (a^x - 1)\frac{1}{\sqrt{a^x}\ln a}
\end{align*}

For the conditions stated $a^x < 1$ so that the left factor is negative. $\sqrt{a^x}$ is positive and $\ln a$ negative so that the right factor is negative as well, giving us a contradiction.
\end{proof}

>>>>

\documentclass[11pt]{article}

@@

The logic behind assumption \ref{box-counting-assumption} is detailed in @!. However, this relation is not guaranteed to hold exactly. A common tactic is to assume the relation 
\begin{align}
N_\epsilon = \Phi\left(\frac{1}{\epsilon}\right)\left (\frac{1}{\epsilon}\right) ^ d
\end{align}

@@

We can change the definition of dimension by introducing a parameter $q$ that determines how much more we care about regions with high probability than about regions with low probability. We define the quantity $I_q$ as an analgoue of the box counting dimension:
\[
I_q(p, \epsilon) = \sum_{c \in C_\epsilon}p(c)^q 
\]
Where $C$ is the covering of the probability distribution in boxes of side length $\epsilon$. The \textit{generalized dimension} of $p$ is then defined as
\[
D_q(p) = \frac{1}{q-1} \lim_{\epsilon \rightarrow 0} \frac{\ln I_q(X, \epsilon)}{\ln \epsilon}
\]

For $D_0$ reduces to the box counting dimension of the support. $D_1$ causes a division by zero, but letting $q \rightarrow 1$ we can formulate 
\[
D_1(p) = lim_{\epsilon \rightarrow 0} \frac{1}{\ln \epsilon} \sum_{c \in C_\epsilon} p(c) \ln p(\epsilon)
\] 
The right part of this is the negative of the entropy of the distribution over $C$ under $p$. In essence $D_1$ measures how the entropy of a covering scales with the accuracy of that covering. This gives a more rigorous background to the intuitive notion of dimension as the amount of number need to describe a point. If we \ldots @! see Theiler.

@@

As noted before, we assume that our data is independently sampled from some probability distribution. This may not be the case. In particular, time series data\ldots

[@! Read Theiler's EEG paper. Discuss natural measures briefly. Also ergodicity?]

@@

The great benefit of the Takens estimator (and other estimators based on the correlation integral) is that it works in distances between data points rather than data points themselves. This has the effect of squaring the volume of available data (barring some reduction of independence).

The drawback is that using all available data can become very expensive, even for small datasets. In the basic algorithm taken from \cite{clauset2007power}, we must sum over all distances to find the estimate for the dimension, then repeat this process to find the estimate for the $d_max$ parameter and finally we must repeat the process around 1000 times to get a reasonably accurate estimate for the significance.

In some contexts this problem can be overcome. If only a single dimension calculation is required after a long process of generating data, then the experimenter will likely not mind waiting a day or so for the results. Alternatively, one could exploit `embarrasingly parallel` nature of the algorithm and spread the search for a fitting $d_max$ or the iterations of the significance calculation over many cores.

If many calculations of dimension are required, or the dataset is so large that even parallelization and patience are insufficient, we can sample the data. The simplest approach of course, is to take a sample of the data points and run the procedure on that sample. This has two drawbacks. First, of we sample from the set of distances instead we will increase the likelihood that our dataset represents a sample of independent distances. More importantly, we are using our sample to estimate two parameters $D_t$ and $d_{\max}$ some crude tests will show very quickly that if we want to estimate $D_t$ as accurately as possible, we should put all our computational power in the evaluation of the maximum likelihood estimator rather than considering many different values for $d_{\max}$ in other words, if we calculate the MLE with all available distances for a single randomly selected value of $d_{\max}$ we are generally better off than when we use to random values for $d_{\max}$ and half the distances for the MLE.

Of course, even if the expected accurary is higher, using a single candidate for $d_{\max}$ is still likely to cause a very high variance. We use the following algorithm as a compromise.

We first generate a set of candidates $C$ with sie $c$ for the value of $d_max$. These candidates are generated by taking a subsample $R'$ of $s$ of the distances and running the full estimator on this subsample. When then add the $d_max$ value of this estimator to list of candidates. Since we can make $s$ relatively big (ie. $5000$) without incurring a great computational cost, this allows us to generate a much better set of candidates then a simple random sample would give us, while not making any further assumptions about the data and our estimator than that more data improves accurary on average. \footnote{It may be of interest to note that if we set $s$ to one, we are essentially sampling $c$ candidates at random for $d_{\max}$ while using all data for the MLE as described before. Thus blind subsampling is a special case of this algorithm.}

It is tempting to look toward search or optimization algorithms over the domain of $d_max$ to minimize the KS statistic, but these will often have rndom behaviours and assumptions on smoothness which will reduce the emprical strength of the resulting answer. Only the complete estimator will guarantee that the returned values will optimize their respective measures (maximum likelihood for $D_t$ and minimum KS statistic for $d_{\max}$). However if we must reduce the size of the observed data, we believe our approach strikes a nice balance between using computational resources efficiently and producing a trustworthy result.

To consider the impact, we test how the accuracy of the estimator and the measure of its significance changes with a reduction in the number of candidates (keeping the number of samples per candidate fixed). The result is shown in figure @! (averging over 100 repeats of the experiment). We see that the $D_t$ estimator is the least affected. The $d_{\max}$ estimator suffers more, but as it is of secondary importance, this shouldn't bother us too much. The significance clearly becomes unreliable at low values of $c$. This is shown not only in the mean value but more importantly in the standard deviation. When using the full estimator, we can expect a good estimate of the significance with low variance, but even small subsampling begins to introduce massive vriance and error. For this reason we will report results on large datasets (using the method detailed here) but we will always take a subsample of the data to compute the significance. We will then make a further assumption that the significance for the subsample is a good estimate for the significance of the estimate on large data.

@@

This is an example of a distribution with multiple intrinsic dimensions. The data is generated by different processes that each create a dimensionality in the data at a particular scale. It's not uncommon to see data with more than two intrinsic dimensions. In the log-log plot produced by the Grassberger-Procaccia estimator, this behavior will produce linear segments for the relevant length scales.

@@

To find the interval limits $r_i$ we proced as we did before with $d_{\max}$ and search for a set of values from the data which minimize the Kolmogorov-Smirnov statistic with the filtered data.

To find multiple, non-overlapping intervals, we use the follwing procedure. We first test all pairs of distances $r_a, r_b$ with $r_a < r_b$ and to find the pair with the lowest KS-statistic. Then, for the ranges to the left and right of $[r_a, r_b)$ we repeat the the procedure on only the data in that range. We repeat the procedure resursively until the data runs out or until the significance for the best subrange of any range drops below 0.01.

@@

Let $R$ be the set of distances in the dataset less than $r_{\max}$. The the likelihood of this dataset is
\[
{\cal L}(R|D_t) = \prod_{r \in R} D_t \frac{1}{r} \left (\frac{r}{r_{\max}} \right ) ^ {D_t} 
\] 
Which we must maximize. Maximizing the logarithm of the likelihood is equivalent, and easier.
\begin{align*}
\ln {\cal L}(R|D_t) &= -\sum \ln r + D_t \sum \ln \frac{r}{r_{\max}} + |R| \ln D_t
\end{align*}

@@

\begin{align*}
\frac{\partial \ln{\cal L}(R|D_t)}{\partial D_t} = \sum{\ln \frac{r}{r_{\max}}} + |R|\frac{1}{D_t}
\end{align*}

@@

\begin{align*}
\sum{\ln \frac{r}{r_{\max}}} + |R|\frac{1}{D_t} &= 0 \\
|R|\frac{1}{D_t} &= - \sum{\ln \frac{r}{r_{\max}}} \\
D_t &= \frac{|R|}{- \sum{\ln \frac{r}{r_{\max}}}} 
\end{align*}

@@

We start with the cumulative distribution function as described in the text. We use $r_{i-1} = r_h$ to clear up the notation.

@@

\begin{align*}
p(d = r \mid \ldots) = {D_i} r^{D_i - 1} r_h^{-D_i} \frac{1}{\left(\frac{r_i}{r_h}\right)^{D_i} - 1}
\end{align*}

This gives us the following log likelihood function:
\begin{align*}
\ln{\cal L}(R_i|D_i) &= \sum \ln D_i +  (D_i-1) \sum \ln r - D_i \sum \ln r_h  - \sum \ln \left(\left(\frac{r_i}{r_h}\right)^{D_i} -1\right)\\
&=  |R_i|\ln D_i + (D_i - 1)\sum \ln r - |R_i| D_i \ln r_h - |R_i| \ln \left(\left(\frac{r_i}{r_h}\right)^{D_i} -1\right) 
\end{align*}

Where $R_i$ is the subset of distances in the dataset that fall within the interval $[r_h, r_i)$.

Taking the derivative:
\begin{align*}
\frac {\partial \ln{\cal L}(R_i|D_i)}{\partial D_i} &=  \frac{|R_i|}{D_i} + \sum \ln r - |R_i|\ln r_h - |R_i| \frac{\left(\frac{r_i}{r_h}\right)^{D_i} \ln \frac{r_i}{r_h}}{\left(\frac{r_i}{r_h}\right)^{D_i} -1}
\end{align*}

@@

\begin{align*}
\frac{1}{D_i}  - \frac{\left(\frac{r_i}{r_h}\right)^{D_i} \ln \frac{r_i}{r_h}}{\left(\frac{r_i}{r_h}\right)^{D_i} -1} &= \frac{|R_i| \ln r_h - \sum \ln r}{|R_i|}  
\end{align*}

We have little hope of solving this for $D_i$, but as shown in lemma \ref{monotonicity}, the function whose root we're looking for is strictly decreasing for the values we are interested in. This allows us to find the value of $D_i$ by binary search as detailed in the main text.

For $r_h = 0$ this equation becomes undefined. However, this is just the single scale case, so we can use the basic Takens estimator derived in the previous section.

@@

@! TODO Prove that it starts at a positive value

@@

\begin{lma}
\label{monotonicity}
The function
\[
f(x) = \frac{1}{x} - \frac{a^x \ln(a)}{a^x - 1} + c
\]
is strictly decreasing for $ x > 0$ and $ a > 1$.
\end{lma}
\begin{proof}
Taking the derivative gives us.

@@

$f$ is stricly decreasing in the given domain if $f'(x) < 0$ holds. Assume for a contradiction that
\begin{align*}
-\frac{1}{x^2} + \left [ \frac{\sqrt{a^x}\ln a}{a^x - 1} \right ]^2 \geq 0
\end{align*}
Which rewrites to 
\begin{align*}
\frac{\sqrt{a^x}\ln a}{a^x - 1}  &\geq \frac{1}{x}\\
\frac{a^x - 1}{\sqrt{a^x}\ln a} &\leq x \\
\frac{a^x}{x\sqrt{a^x}\ln a} - \frac{1}{x\sqrt{a^x}\ln a} &\leq 0 \\
a^x &\leq 1
\end{align*}

For $a>1$ and $x>0$ we know that $a^x > 1$ so we have a contradiction.
\end{proof}

\subsection{The Kolmogorov Smirnov test as an MDL approximation}
